A protocol and event-sourced database for decentralized user-siloed data

### 2019-09-27 draft.tex
hyp
Signed-off-by: andrew <andrew@textile.io>
title fix
Signed-off-by: andrew <andrew@textile.io>
review: incremental updates
Signed-off-by: Carson Farmer <carson.farmer@gmail.com>

\title[A protocol and event-sourced database for decentralized user-siloed data]{A protocol and event-sourced database for decentralized user-siloed data}
\author{Pick}
\author{Farmer}
\author{Sutula}
\affiliation{www.textile.io} \email{research@textile.io}
\author{?Gozalishivili others?}
\affiliation{other?} \email{research@textile.io}
\author{Hill}
\affiliation{www.textile.io} \email{contact@textile.io}

@@

Compared to their predecessors, modern cloud-based apps and services provide an extremely high level of convenience. Users can entirely forget about the risk of data loss and enjoy seamless access to their apps across multiple devices. This convenience is now expected, but has come at the cost of additional consequences for users. One such consequence is that many of the same development patterns that bring convenience (e.g., single sign-on, minimal encryption, centralized servers, and databases) also enable or even require data hoarding by the apps. While collecting large amounts of users' data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction; Apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it. Additionally,  the data may not be readily accessible to other apps and services. This is app-siloed data.

While companies collecting user data may not be itself a significant problem, it has been shown that over time company incentives often shift from providing value to users to extracting value from users \cite{dixonWhyDecentralizationMatters2018}. When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value or even revenue for the company. It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data. Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data. This approach is called user-siloed data and it fundamentally separates apps from their user's data.

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app silo-ed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}. However, none so far have provided a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data. This is the key to building great applications on user-siloed data.

In this paper, we study existing technologies that could be used to build a network for user-siloed data. We outline six challenge areas: flexible data format, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication. Based on our findings, we propose a novel architecture for event sourcing with Interplanetary Linked Data (IPLD), designed to store, share, and host user-sioled datasets at scale. Our proposed design leverages new and existing protocols to solve major challenges involved with building a secure and distributed network for user data while at the same time providing the flexibility and scalability required by today's applications.

\section{Background}
\label{sec:Background}

We now describe some of the technologies and concepts motivating the design of our novel decentralized database system. We highlight some of the advantages and lessons that can be learned from event-sourcing and discuss drawbacks to using these approaches in decentralized systems. We provide an overview of some important technologies related to IPFS that make it possible to rethink event sourcing in a decentralized network. Finally, we cover challenges of security and access control on an open and decentralized network and discuss how they are used in popular databases external to IPFS.

@@

@@

When developing large-scale software systems, it is common to store data in a relational database management system (RDBMS). To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database. A powerful alternative is to use a set of append only logs to model the state of an object simply by applying its change sequence in the correct order. This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state. This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{jaykrepsLogWhatEvery2013}.

In modern applications it is critical to have reliable mechanisms for publishing updates and events (i.e., to support event-driven architectures), scalability (optimized write and read operations), forward-compatible application updates (e.g., code changes, retroactive events, and others), auditing systems, etc. To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append only logs to support immutable state histories. Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate event sourcing and CQRS-based applications, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others (see \cite{kleppmannDesigningDataintensiveApplications2017}).

\begin{definition} (Command query responsibility segregation). Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data \cite{martinfowlerCQRS2011}. \end{definition}

\begin{definition} (Event sourcing). Event sourcing (ES) is a design pattern for persisting the state of an application as an append-only log of state-changing events. \end{definition}

ES is particularly useful in contexts where system components are distributed or decentralized, such as in decentralized applications (DApps). This is because the design of systems that use event sourcing often necessitate infrastructure for brokering messages between loosely coupled software components and services. This feature will become increasingly important as we outline background concepts related to data synchronization in the following sections.

@@

\begin{enumerate}
	\item Relational
    \begin{enumerate}
        \item MariaDB \url{https://mariadb.org/}
    	\item PostgreSQL \url{https://www.postgresql.org/}
        \item SQLite \url{https://www.sqlite.org}
    \end{enumerate}
    \item Key value stores
    \begin{enumerate}
        \item Dynamo \url{https://aws.amazon.com/dynamodb/}
    	\item LevelDB \url{https://github.com/google/leveldb}
        \item Redis \url{https://redis.io/}
    \end{enumerate}
    \item Document stores
    \begin{enumerate}
        \item CouchDB \url{http://couchdb.apache.org/}
    	\item MongoDB \url{https://www.mongodb.com/}
        \item RethinkDB \url{https://rethinkdb.com/}
    \end{enumerate}
\end{enumerate}

>>>>

\title[Textile Threads]{A protocol and event-sourced database for decentralized user-siloed data}
\author{Pick}
\author{Farmer}
\author{Sutula}
\affiliation{www.textile.io} \email{research@textile.io}
\author{?Gozalishivili others?}
\affiliation{other?} \email{research@textile.io}
\author{Hill}
\affiliation{www.textile.io} \email{contact@textile.io}

@@

Compared to their predecessors, modern cloud-based apps and services provide an extremely high level of convenience. Users can completely forget about data management, and enjoy seamless access to their apps across multiple devices. This convenience is now expected, but has come at the cost of additional consequences for users. One such consequence is that many of the same development patterns that bring convenience (e.g., single sign-on, minimal encryption, centralized servers and databases) also enable, or even require data hoarding by apps. While collecting large amounts of users' data can create value for the companies building apps (e.g., app telemetry, predictive tools, or even new revenue streams), that value flows mostly in one direction: apps collect, process and benefit from a user's private data, but users rarely have access to the original data or the new insights that come from it. Additionally, the data is generally not accessible to other apps and services. This is \emph{app-siloed} data.

While companies collecting user data may not itself be a significant problem, it has been shown that, over time company incentives often shift from \emph{providing} value to users, to \emph{extracting} value from users \cite{dixonWhyDecentralizationMatters2018}. When this incentive shift happens for companies that have also been hoarding user data, that data may be a new source of value/revenue for the company. It may be possible to stop this trend through extreme privacy measures, government intervention/legislation, or by boycotting companies that collect any form of data. Ideally, there is an alternative approach that allows individuals to capture the value of their data and still allow developers to build new interfaces and experience on top of this data. This approach is called \emph{user-siloed} data and it fundamentally separates apps from their users' data.

One of the most exciting aspects of user-siloed data is the the ability to build data-driven apps and services while users remain in control of their own data. Other projects have identified app-siloed data as a problem \cite{bernersLee_readwrite_2013,de_montjoye_trusted_2012}, and some have identified user-siloed data as a solution \cite{sambraSolidPlatformDecentralized2016}. However, none so far have addressed the problem directly, by providing a sufficiently interoperable protocol for scalable storage and transmission of user-siloed application data.

In this paper, we study existing technologies that could be used to build a network for user-siloed data. We outline six challenge areas: flexible data formats, efficient synchronization, conflict resolution, access-control, scalable storage, and network communication. Based on our findings, we propose a novel architecture for event sourcing (ES) with Interplanetary Linked Data (IPLD), that is designed to store, share, and host user-siloed datasets at scale. Our proposed design leverages new and existing protocols to solve major challenges with building a secure and distributed network for user data while at the same time providing the flexibility and scalability required by today's apps.

\section{Background}\label{sec:Background}

We now describe some of the technologies and concepts motivating the design of our novel decentralized database system. We highlight some of the advantages and lessons learned from event sourcing (ES) and discuss drawbacks to using these approaches in decentralized systems. We provide an overview of some important technologies related to the Interplanetary File System (IPFS) that make it possible to rethink ES in a decentralized network. Finally, we cover challenges to security and access control on an open and decentralized network, and discuss how they are used in popular databases outside the IPFS context.

@@

When developing large-scale software systems, it is common to store data in a relational database management system (RDBMS). To model realistic systems, this type of framework often requires complex techniques for mapping data between domain models and database tables, where the same data model is used to both query and update a database. A powerful alternative is to use a set of append-only logs to model the state of an object simply by applying its change sequence in the correct order. This concept can be expressed succinctly by the state machine approach \cite{schneiderImplementingFaulttolerantServices1990}: if two identical, deterministic processes begin in the same state and get the same inputs in the same order, they will produce the same output and end in the same state. This is a powerful concept baked into a simple structure, and is at the heart of many distributed database systems \cite{jaykrepsLogWhatEvery2013}.

@@

For most apps, it is critical to have reliable mechanisms for publishing updates and events (i.e., to support event-driven architectures), scalability (optimized write and read operations), forward-compatible application updates (e.g., code changes, retroactive events), auditing systems, etc. To support such requirements, developers have begun to utilize event sourcing and command query responsibility segregation (CQRS) patterns \cite{bettsExploringCQRSEvent2013}, relying on append-only logs to support immutable state histories. Indeed, a number of commercial and open source software projects have emerged in recent years that facilitate ES and CQRS-based designs, including Event Store \cite{EventStore}, Apache Kafaka \cite{ApacheKafka} and Samza \cite{ApacheSamza}, among others \cite{kleppmannDesigningDataintensiveApplications2017}.

\begin{definition}
  (CQRS). Command query responsibility segregation or CQRS is a design pattern whereby reads and writes are separated into different models, using commands to write data, and queries to read data \cite{martinfowlerCQRS2011}.
\end{definition}

\begin{definition}
  (ES). Event sourcing or ES is a design pattern for persisting the state of an application as an append-only log of state-changing events.
\end{definition}

@@

\begin{enumerate}
	\item Relational
    \begin{enumerate}
        \item MariaDB \url{https://mariadb.org/}
    	\item PostgreSQL \url{https://www.postgresql.org/}
        \item SQLite \url{https://www.sqlite.org}
    \end{enumerate}
    \item Key-value stores
    \begin{enumerate}
        \item Dynamo \url{https://aws.amazon.com/dynamodb/}
    	\item LevelDB \url{https://github.com/google/leveldb}
        \item Redis \url{https://redis.io/}
    \end{enumerate}
    \item Document stores
    \begin{enumerate}
        \item CouchDB \url{http://couchdb.apache.org/}
    	\item MongoDB \url{https://www.mongodb.com/}
        \item RethinkDB \url{https://rethinkdb.com/}
    \end{enumerate}
\end{enumerate}
